---
title: "Simple Example for Geospatial Visualization"
author: "Juliane Manitz"
date: "`r format(Sys.time(),  '%d %B, %Y')`"
output: 
  html_document: 
    self_contained: yes
---

<!-- ## Libraries and settings -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=10)

library(sf)
library(dplyr)
library(ggplot2)
library(viridis)
```

## German District Polygon map

Source: https://sgx.geodatenzentrum.de/wfs_vg5000_1231; Deutsche Verwaltungsgebiete VG5000

```{r}
dtl <- st_read("data/vg5000_ebenen_0101/VG5000_KRS.shp")

# Centered around Berlin EPSG:3068 (just for fun, because I grew up there)
dtl_proj <- st_transform(dtl, crs = 3068)
plot(dtl_proj["geometry"], axes = TRUE, graticule = st_crs(dtl))

# There is not much data in the data, but we can color-code the different states
ggplot(data = dtl_proj) +  
  geom_sf(aes(fill = SN_L)) +
  theme_minimal() 
```

## Agricultural businesses in Germany 1999 - 2007; 

Source: https://www.regionalstatistik.de; table 41120-01-02-4; Landwirtschaftliche Betriebe und landwirtschaftlich genutzte Fläche (LF) n. Kulturarten - Erhebungsjahr - regionale Tiefe: Kreise und krfr. Städte

```{r}
dt <- read.csv("data/41120-01-02-4.csv", sep=";", skip=9, na.strings = c("-","."), 
           col.names=c("year", "ARS", "name", "n_business", "ha_business", "n_farm", "ha_farm", "n_pcrop", "ha_pcrop", "n_grass", "ha_grass")) %>% 
  filter(year != ",,") %>% 
  # Factor handling 
  mutate_if(is.character, factor, exclude = c("", NA)) 
```

## Visualization

### Plot all agricultural data together

In addition to the cities if Berlin and Hamburg, it looks like there are missings from two states in the north-east and east.

```{r}
dtl07 <- dt %>% 
  # use 2007 data only
  filter(year == "2007") %>% 
  # match to map data
  left_join(dtl, ., by ="ARS") 

plot(dtl07[c("n_business", "ha_business", "n_farm", "ha_farm", "n_pcrop", "ha_pcrop", "n_grass", "ha_grass")], 
     logz = TRUE, border = "darkgray")
```

### Plot farmland area over time

No obvious changes over time, probably requires appropriate normalization

```{r}
plot_farmland <- function(i){

  # subset data 
  dt %>% filter(year == i) %>% 
    # match to map data
    left_join(dtl, ., by ="ARS") %>% 
    # plot map
    ggplot(data = .) +  
    geom_sf(aes(fill = ha_farm/1000)) + scale_x_sqrt() +
    labs(title = paste0("Farmland area in ",i)) + 
    scale_fill_viridis() + theme_minimal()
}

pp <- lapply(levels(dt$year), plot_farmland)
ggpubr::ggarrange(plotlist=pp, legend= "right", common.legend = TRUE)
```

# Session Information (for Reproducibility)

```{r}
print(sessionInfo(), locale = FALSE)
```
