---
title: "Temperature Shift in Boston"
author: "Juliane Manitz"
date: "`r format(Sys.time(),  '%d %B, %Y')`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(readxl)
library(tidyr)
require(purrr)
require(magrittr)

require(gridExtra)
library(ggplot2)
library(viridis)
```


## Load Data

NOWData - NOAA Online Weather Data. Monthly summarized data calculates averages, totals, daily extremes or frequencies for the selected variable for each month of the year for the selected range of years. Note: trace precipition/snowfall/snow depth amounts are treated as zero in sums, mean, and number of days counts. Annual average temperatures are the average of the twelve monthly values. Temperatures are reported in degrees F; precipitation, snowfall and snow depth are reported in inches. Source: https://www.weather.gov/wrh/Climate?wfo=box

```{r}
# To load all sheets in a workbook, use lapply()
path <- "Temp_Boston.xlsx"
dt_list <- map(excel_sheets(path), read_excel, path = path, skip=2)
names(dt_list) <- excel_sheets(path)
#dt_list <- lapply(excel_sheets(path), read_excel, path = path, skip=2)

# Data preprocessing
dt_list2 <- lapply(names(dt_list), function(x)
  dt_list[[x]] %>%  
    mutate_if(is.character, list(~na_if(., "M") %>% na_if(., "T") %>% as.numeric)) %>%
    pivot_longer(-Year, names_to = "Month", values_to = x) #  assign(x, df_list[[x]], envir = .GlobalEnv)
)

# Combine list into one dataset
dt <- Reduce(left_join, dt_list2)

# Separate Annual averages from monthly values
dtY <- dt %>% filter(Month == "Annual")
dtM <- dt %>% filter(Month != "Annual") %>% 
  mutate(date = paste0("01-",tolower(Month),"-",Year) %>% as.Date(format = "%d-%b-%Y"))
```

## Time Series 

```{r}
dtY %>% filter(Year<1950) %>% summarize(mean(meanTemp))
ggplot(dtY, aes(x=Year, y=meanTemp)) +
  geom_line() + stat_smooth(method = "loess", se=FALSE, formula = y ~ x, size = 1) +
  geom_hline(yintercept = dtY %>% filter(Year<1950) %>% summarize(mean(meanTemp)) %>% pull(), 
             linetype = 2, color = "darkred", linewidth = 1) + 
  scale_x_continuous(breaks=seq(1850, 2020, 10)) + 
  labs(x="year", y="Annual Temperature (in F)", title="Annual Average Temperature in Boston, MA") +
  theme_minimal()
```

## Summer Temperature Shift 

```{r}
# Define bins for coloring based on early data < 1950
cols <-c("cornflowerblue","cornsilk","brown1","darkred") 
cc <- dtM %>% filter(Month %in% c("Jun","Jul","Aug","Sep") & Year < 1950) %>% 
  pull(meanTemp) %>% quantile(probs =c(.325,.5,.95))
dtM %<>% mutate(meanTempC = factor(cut(meanTemp, breaks = c(0,cc,120)), 
                labels=c("Cooler than average", "Average", "Warmer than average", "Extremely hot")))
```

```{r, fig.height=10, fig.width=6, warning=FALSE}
p1 <- dtM %>% 
  filter(Month %in% c("Jun","Jul","Aug", "Sep") & Year < 1950) %>% 
  ggplot(aes(meanTemp, fill = meanTempC)) + scale_fill_manual(values = cols) + 
  geom_histogram(bins = 27, show.legend = TRUE) +
  labs(y="Frequency", x="Temperature (in F)", 
       title="Summer Temperatures in Boston, MA", subtitle = "1872-1950", fill="") +
  scale_x_continuous(limits = c(57,82), breaks=seq(50,120, 5)) + 
  theme_minimal() + theme(legend.position = c(0.85,0.75))

p2 <- dtM %>% 
  filter(Month %in% c("Jun","Jul","Aug", "Sep") & Year < 1990 & Year > 1950) %>% 
  ggplot(aes(meanTemp, fill = meanTempC)) + scale_fill_manual(values = cols) + 
  geom_histogram(bins = 27, show.legend = FALSE) +
  labs(y="Frequency", x="Temperature (in F)", subtitle = "1950-1990", fill="") +
  scale_x_continuous(limits = c(57,82), breaks=seq(50,120, 5)) + 
  theme_minimal() 

p3 <- dtM %>% 
  filter(Month %in% c("May","Jun","Jul","Aug", "Sep") & Year > 1990) %>% 
  ggplot(aes(meanTemp, fill = meanTempC)) + scale_fill_manual(values = cols) + 
  geom_histogram(bins = 27, show.legend = FALSE) +
  labs(y="Frequency", x="Temperature (in F)", subtitle="1990-2024", fill="") +
  scale_x_continuous(limits = c(57,82), breaks=seq(50,120, 5)) + 
  theme_minimal() 

grid.arrange(p1,p2,p3,ncol=1)
```

## Session Information (for Reproducibility)

```{r}
print(sessionInfo(), locale = FALSE)
```